---
title: "DATA 643 Recommender System"
author: "Puneet Auluck"
date: "February 13, 2017"
output: html_document
---

This recommender system will use MovieLense data from `recommenderlab` package.  This system will recommend top 5 movies to randomly selected users using `item-based collaborative filtering`.


```{r, echo=FALSE}
library(recommenderlab)
library(dplyr)
library(lsa)
library(knitr)
```

### Load data

```{r}
data("MovieLense")
movies <- as.matrix(MovieLense@data)

dim(movies)

```

This data contains 943 users with ratings on 1664 movies.

### Normalize data

The dataset will be normalized taking standard score using universal mean.

```{r}
movies[movies==0] <- NA
mmean <- mean(movies, na.rm = T)
msd <- sd(movies, na.rm=T)
moviesnorm <- (movies - mmean)/msd

```

### Generate similarity matrix 

We will be generating similarity matrix based on similarities between items.

```{r}
library(lsa)
movietemp <- movies
movietemp[is.na(movietemp)] = 0
movie_sim_mtx = cosine(as.matrix(movietemp))

#image(as.matrix(movie_sim_mtx))
#library(corrplot)
#corrplot(as.matrix(movie_sim_mtx), is.corr = FALSE, method = "color")
```

### Create function to return top 5 recommendations for a user

Populate all `na` values for a user 

```{r}

getTopFive <- function(u){
  
  # get all normalized ratings for this user
  user_ratings <- moviesnorm[u,]

  # get list of movies user did not rate
  movies_user_not_rated <- which(is.na(user_ratings))

  # get list of movies user rated
  movies_user_rated <- which(!is.na(user_ratings))
  
  # list of movie names 
  movie_names <- names(user_ratings)
  
  # list to store predicted values for not rated moves
  now_rated_value <- list()
  now_rated_name <- list()

  # calculated weighted similarity for non related
  for(i in 1:length(movies_user_not_rated)){
    # get column from similarity matrix
    movie_item_simlarity = movie_sim_mtx[i,]
    
    # calculate weight and weight average
    weight_total = sum(movie_item_simlarity[unlist(movies_user_rated)])
    weighted <- movie_item_simlarity * moviesnorm[1,]
    
    # store the movie rating and its name
    now_rated_value[i] <- sum(weighted,na.rm=T)/weight_total
    now_rated_name[i] <- movie_names[movies_user_not_rated[i]]
  }

  # return top five recommended values
  now_rated_df <- cbind.data.frame("m"=unlist(now_rated_name),
                                   "v"=unlist(now_rated_value), 
                                   stringsAsFactors = FALSE)
  now_rated_sorted_df <- now_rated_df[order(-now_rated_df$v),] 
  rv <- head(now_rated_sorted_df$m,n=5)

  return(rv)
}


```

### Randomly pick 10 users to display their top 5 movie picks

```{r}
# Print out top 5 recommendations for 10 users
recommends <- data.frame(matrix(nrow=10,ncol=6))
names(recommends) <- c("User","Top1","Top3","Top3","Top4","Top5")

# randomly pick 10 users
set.seed(123)
users <- sample(1:nrow(moviesnorm),10, replace = FALSE)

for(i in c(1:length(users))){
  top5 <- getTopFive(users[i])
  recommends[i,] <- c(users[i],as.list(top5))
}

# print out the recommendations
kable(recommends)
```


### Recommendation system with built in system

```{r}
# collect movie names and its index
movie_names <- colnames(movies)

# run recommenderlab builtin function
recsys <- Recommender(data = MovieLense, method = "IBCF", parameter = list(method = "Cosine"))
recsysTop5 <- predict(object = recsys, newdata = MovieLense, n = 5)
top5items <- recsysTop5@items

# get top 5 movie names
top5builtin <- movie_names[unlist(top5items[22])]

# Print out top 5 recommendations for 10 users
recommends_builtin <- data.frame(matrix(nrow=10,ncol=6))
names(recommends_builtin) <- c("User","Top1","Top3","Top3","Top4","Top5")

for(i in c(1:length(users))){
  top5 <- movie_names[unlist(top5items[users[i]])]
  recommends_builtin[i,] <- c(users[i],as.list(top5))
}

kable(recommends_builtin)
```

From running my own recommender algorithm and built-in, the built-in processes the feed much faster.  Unfortunately, the outputs from both systems do not match which tells me that should improve on the system I have created.

***