---
title: "DATA 643 Recommender System - Project #3"
author: "Puneet Auluck"
date: "March 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This recommender system will use MovieLense data from `recommenderlab` package. 


```{r, echo=FALSE, message=FALSE}
library(recommenderlab)
library(dplyr)
library(irlba)
library(ggplot2)
```

#### Load data

```{r}
data("MovieLense")
movies <- as.matrix(MovieLense@data)
```


#### Let's explore the data
```{r}
dim(movies)

```


#### Check sparsity of the data

```{r}
length(movies[movies==0]) / (ncol(movies)*nrow(movies))
```

The matrix is 93% sparse, meaning only 7% of entries are populated with ratings.


#### Find universal mean

The dataset will be normalized taking standard score using universal mean after SVD matrix is created.

```{r}

moviesWithNA <- movies
is.na(moviesWithNA) <- moviesWithNA == 0
universalMean <- mean(moviesWithNA, na.rm=T)

```



#### Use SVD technique to decompose movies matrix

Our goal will be to reduce `movies` matrix and eliminate the less important attributes to make accurate predictions. 

Singular value decomposition is defined as

$M = U  \Sigma  V^T$

where M is `mxn`, U is `mxr`, $\Sigma$ is `rxr` and $V^T$ is `rxn`.

When we set rank to r, it is full rank but when we set rank to k where `k<r`, we get reduced matrix:

$A_k = U_k  S_k  V_k^T$

The item profile is defined as: $U_k\sqrt{\Sigma_k}$

The user profile is defined as: $\sqrt{\Sigma_k}V_k^T$

The item-user profile is defined as: $U_k\sqrt{\Sigma_k}\sqrt{\Sigma_k}V_k^T$

Also, to determine how well the predictions were made, we define RMSE as

RMSE = $\sqrt{\frac{1}{N} \Sigma (r_i - \hat{r_i})^2}$

```{r}

RMSE <- function(svdmtx){
  N <- nrow(svdmtx)*ncol(svdmtx)
  sse <- sum((svdmtx - movies)^2)
  rv <- sqrt(sse)/N
  return(rv)
}
  
```

Let us begin by setting k= 100 and plot it.

```{r}
decomp <- irlba(movies, nv=100)
results <- cbind.data.frame(c(1:100), decomp$d)
colnames(results) <- c("rank","singular_value")
ggplot(results, aes(rank, singular_value) )+ geom_point()
#plot(c(1:100),decomp$d)

```

It appears most imporant features are captured within 35 singular values.  Let's run SVD with k=35.

```{r}

decomp35 <- irlba(movies, nv=35)
sigma <- decomp35$d
results <- cbind.data.frame(c(1:35), decomp35$d)
colnames(results) <- c("rank","singular_value")
ggplot(results, aes(rank, singular_value) )+ geom_point()

```


Calculate the item-profile matrix

```{r}

movieSVD <- universalMean + (decomp35$u %*% sqrt(sigma)) %*% (sqrt(sigma) %*% t(decomp35$v))
colnames(movieSVD) <- colnames(movies)
rownames(movieSVD) <- rownames(movies)

RMSE(movieSVD)

decomp20 <- irlba(movies, nv=20)
sigma <- decomp20$d

movieSVD <- universalMean + (decomp20$u %*% sqrt(sigma)) %*% (sqrt(sigma) %*% t(decomp20$v))
colnames(movieSVD) <- colnames(movies)
rownames(movieSVD) <- rownames(movies)

RMSE(movieSVD)


```

#### Evaluate Built-in IBCF and UBCF Recommenders

Let us evaluate and compare RMSE values with built-in recommender methods

```{r}
moviesEval <- as(movies, "realRatingMatrix")
eval_sets <- evaluationScheme(data = moviesEval, method = "split", train = 0.9, given = 20, goodRating = 3, k = 1)

#IBCF
eval_recommender <- Recommender(data = getData(eval_sets, "train"), method = "IBCF", parameter = NULL)
eval_prediction <- predict(object = eval_recommender, newdata = getData(eval_sets, "known"), n = 10, type = "ratings")
calcPredictionAccuracy(x = eval_prediction, data = getData(eval_sets, "unknown"), byUser = FALSE)

#UBCF
eval_recommender <- Recommender(data = getData(eval_sets, "train"), method = "UBCF", parameter = NULL)
eval_prediction <- predict(object = eval_recommender, newdata = getData(eval_sets, "known"), n = 10, type = "ratings")
calcPredictionAccuracy(x = eval_prediction, data = getData(eval_sets, "unknown"), byUser = FALSE)



```

The prediction created from SVD appears to be the least, followed by UBCF method.  The IBCF method used from 'recommenderlab` produced the largest RMSE value.

***